\large Neural, Tree, and Symbolic Optimization of Cosmological Particle-Mesh Simulation \vspace{0.3em}

### 2023-06-25 docs/papers/sto/so.tex
FIXUP draft: clean up the recap on adjoint

The differentiable cosmological simulation with adjoint method \citep{Li2022a}
and the corresponding PM $N$-body library \pmwd\ \citep{Li2022b} are the
foundation of the optimization discussed in this work.
We briefly recap the discussion in \citet{Li2022a} and introduce the notations.
We denote the particles' state with $\vz = (\vx, \vp)^\intercal$ and their
adjoint variables with $\vlambda = (\vxi, \vpi)$.
The forward simulation is evolved using a symplectic integrator.
From time step $a_{i-1}$ to $a_i$, the particles' state is updated alternately
for $\vx$ and $\vp$ through a loop over the symplectic integration indexed with
$j$
\begin{alignat}{2}
&D_{j-1}^j: &\qquad \vx_j &= \vx_{j-1} +
  \vp_{j-1} D(a^p_{j-1}, a^x_{j-1}, a^x_j), \nonumber\\
%
&F_j: & \va_j &\triangleq -\vnabla\varphi(\vx_j), \nonumber\\
%
&K_{j-1}^j: & \vp_j &= \vp_{j-1} + \va_j K(a^x_j, a^p_{j-1}, a^p_j).
\label{DFK}
\end{alignat}
The intermediate times are given by
\begin{align}
a^x_j &= (1 - C^x_j) a_{i-1} + C^x_j a_i, \nonumber\\
a^p_j &= (1 - C^p_j) a_{i-1} + C^p_j a_i,
\end{align}
where $C^x$ and $C^p$ are the cumulative coefficients of the symplectic
integrator, e.g. $C^x = (0, 0, 1)$ and $C^p = (0, 0.5, 1)$ for leapfrog.
The reverse time integration from $a_i$ to $a_{i-1}$ and adjoint equations
\begin{alignat}{2}
&K_j^{j-1}: &\qquad \vp_{j-1} &= \vp_j +
  \va_j K(a^x_j, a^p_j, a^p_{j-1}), \nonumber\\
%
& & \vxi_{j-1} &= \vxi_j - \valpha_j K(a^x_j, a^p_j, a^p_{j-1}) \nonumber\\
%
&D_j^{j-1}: & \vx_{j-1} &= \vx_j +
  \vp_{j-1} D(a^p_{j-1}, a^x_j, a^x_{j-1}), \nonumber\\
%
& & \vpi_{j-1} &= \vpi_j - \vxi_{j-1} D(a^p_{j-1}, a^x_j, a^x_{j-1}), \nonumber\\
%
&F_{j-1}: & \va_{j-1} &\triangleq -\vnabla\vvarphi(\vx_{j-1}), \nonumber\\
%
& & \valpha_{j-1} &\triangleq -\vpi_{j-1} \cdot
  \frac{\p\vnabla\vvarphi_{j-1}}{\p\vx_{j-1}}, \nonumber\\
& & \vzeta_{j-1} &\triangleq - \vpi_{j-1} \cdot
  \frac{\p\vnabla\vvarphi_{j-1}}{\p\vtheta}.
\label{KDF_adj}
\end{alignat}

We consider an objective function $\cJ(\vo, \theta)$ defined on the observable
$\vo(\vz_0,...,\vz_n,\vtheta)$.
The observe operator from $a_i$ to $a_{i-1}$ reads
\begin{align}
  O_i^{i-1}: \qquad \vxi_{i-1} :\:\!\!\!&= \vxi_{i-1}
  + \frac{\p\cJ}{\p\vo} \cdot \frac{\p\vo}{\p\vx_{i-1}}, \nonumber\\
                    \vpi_{i-1} :\:\!\!\!&= \vpi_{i-1}
  + \frac{\p\cJ}{\p\vo} \cdot \frac{\p\vo}{\p\vp_{i-1}}.
\end{align}

% % TODO update the final gradient for loop over j
% The final gradient is accumulated through
% \begin{multline}
%   \frac{\d\cJ}{\d\vtheta} = \frac{\p\cJ}{\p\vtheta}
%   %
%   + \frac{\p\cJ}{\p\vo} \cdot \frac{\p\vo}{\p\vtheta}
%   %
%   + \vxi_0 \cdot \frac{\p\vx_0}{\p\vtheta}
%   + \vpi_0 \cdot \frac{\p\vp_0}{\p\vtheta} \\
%   %
%   - \sum_{i=1}^n \Bigl[
%     \bigl( \vpi_i \cdot\va_i \frac\p{\p\vtheta} + \vzeta_i
%       \bigr) K(t_i, t_i, t_{i-\half}) \\
%     + \vxi_i \cdot \vp_{i-\half}
%       \frac{\p D(t_{i-\half}, t_i, t_{i-1})}{\p\vtheta}
%     \Bigr],
% \label{KDFO_grad}
% \end{multline}
% where $\vtheta$ could include cosmological parameters, initial conditions, as
% well as neural network parameters for optimization in this work.

>>>>

The differentiable cosmological simulation with adjoint method \citep{Li2022a}
and the corresponding PM $N$-body library \pmwd\ \citep{Li2022b} are the
foundation of the optimization discussed in this work.
We denote the particles' state with $\vz = (\vx, \vp)^\intercal$ and their
adjoint variables with $\vlambda = (\vxi, \vpi)$.
